Supplementary material for  \nBayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient

### 2023-02-02 manuscript/manuscript.Rmd
results and intro updates

Abstract
===============================================================================
Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time. However, limited spatial coverage and disagreements between geochemical proxies compromise the accuracy of these reconstructions. Geological data such as coral reefs, mangroves, and evaporites have the potential to supplement geochemical data by improving spatial coverage and constraining temperature estimates. Historically, these data have been used to qualitatively inform upon palaeoclimatic conditions at broad spatial scales, yet no transfer functions exist to convert geological data into temperature estimates. Quantitative temperature reconstructions have therefore not made use of these data. Here, we present a Bayesian hierarchical model to integrate geological data–with established temperature proxies–into a unified quantitative framework. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO. We show that the inclusion of ecological data can substantially reduce the uncertainty on temperature estimates in climate zones lacking geochemical data.

# Introduction
The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021]. Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018]. Reconstructions of palaeoclimate commonly use geochemical proxy data, palaeobiological or lithological climate indicators, or earth system modelling.

@@

## Parameter estimation
We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations. The modern temperature gradient with the Eocene sampling distribution was estimated in one chain with 25,000 iterations for each of the 100 temperature samples. 5,000 iterations each were discarded as burn-in, and every 10th iteration was kept, resulting in a total of 200,000 iterations across all 100 model runs.
For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000). Trace plots were used to assess the convergence and mixing of the MCMC chains (Supplementary Materials S1).

@@

@@

# global average temperatures
motemp_global <- readRDS("results/modern/global_mean_modern.rds")
eotemp_global <- readRDS("results/eeco/global_mean_eeco.rds")
moeotemp_global <- readRDS("results/eeco/global_mean_modern_eocene_sampling_eeco.rds")
moeotemp_global_diff <- readRDS("results/eeco/global_mean_modern_eocene_sampling_difference.rds")
eoceneproxy_global_diff <- readRDS("results/eeco/global_mean_eocene_eoceneproxy_difference.rds")

@@

Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures. The modern, global mean sea surface temperature estimate is `r motemp_global$"50%"` degC when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern global mean sea surface temperature is estimated at `r moeotemp_global$"50%"` (95% CI: `r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.

@@

The modeled Eocene temperature gradient is starkly different from the modern (Fig 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (95% CI: `r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (95% CI: `r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.

The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`. This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The global early Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (95% CI: `r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).

A model run excluding the ecological proxies results in a similar median latitudinal gradient (`r eocene_proxy_gradient$"50%"` degC), but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxy data, while providing additional constraints on the shape of the early Eocene temperature gradient.

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model seperately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by ...

>>>>

Abstract
===============================================================================
Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time. However, limited spatial coverage and disagreements between geochemical proxies compromise the accuracy of these reconstructions. Geological data such as coral reefs, mangroves, and evaporites have the potential to supplement geochemical data by improving spatial coverage and constraining temperature estimates. Historically, these data have been used to qualitatively inform upon palaeoclimatic conditions at broad spatial scales, yet no transfer functions exist to convert geological data into temperature estimates. Quantitative temperature reconstructions have therefore not made use of these data. Likewise, the large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into consideration the spatial distribution of the samples. Here, we present a Bayesian hierarchical model to integrate geological data–with established temperature proxies–into a unified quantitative framework. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO. We show that the inclusion of ecological data can substantially reduce the uncertainty on temperature estimates in climate zones lacking geochemical data.

# Introduction
The increasing availability of compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g. @Veizer2015; @Hollis2019; @Grossman2022; @Judd2022]. In order for these reconstructions to be effective, two fundamental challenges need to be overcome: 1) disparate types of proxy data need to be integrated to inform a single climate reconstruction, and 2) large-scale scale climate patterns need to be inferred from localised, often sparse observations.

For the marine realm, most palaeoclimate reconstructions use geochemical proxy data

The first challenge requires that all proxy data is converted to a common scale.

A fundamental challenge in reconstructing Earth's past climate is to infer large-scale climatic patterns from local observations, derived from disparate proxies.  In order for the

Proxy-derived palaeoclimate reconstructions offer an important alternative to and a test of Earth system models, and as such are fundamental for reconstructing Earth's past climate

The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021]. Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018]. Reconstructions of palaeoclimate commonly use geochemical proxy data, palaeobiological or lithological climate indicators, or earth system modelling.

@@

## Parameter estimation
We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations. The modern temperature gradient with the Eocene sampling distribution was estimated in one chain with 25,000 iterations for each of the 100 temperature samples. 5,000 iterations each were discarded as burn-in, and every 10th iteration was kept, resulting in a total of 200,000 iterations across all 100 model runs.
For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).

@@

eocene_north_south_gradient_diff <- readRDS("./results/SM/eocene_north_south_gradient_diff.rds")

@@

# global average temperatures
motemp_global <- readRDS("results/modern/global_mean_modern.rds")
eotemp_global <- readRDS("results/eeco/global_mean_eeco.rds")
moeotemp_global <- readRDS("results/eeco/global_mean_modern_eocene_sampling_eeco.rds")
moeotemp_global_diff <- readRDS("results/eeco/global_mean_modern_eocene_sampling_difference.rds")
eoceneproxy_global_diff <- readRDS("results/eeco/global_mean_eocene_eoceneproxy_difference.rds")
eeco_north_global <- readRDS("results/SM/global_mean_northern_eeco.rds")
eeco_south_global <- readRDS("results/SM/global_mean_southern_eeco.rds")
eeco_south_north_global_diff <- readRDS("results/SM/global_mean_diff_south_north_eeco.rds")

@@

Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures. The modern, global mean sea surface temperature estimate is `r motemp_global$"50%"` degC when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern global mean sea surface temperature is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.

@@

The modeled Eocene temperature gradient is starkly different from the modern (Fig 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.

The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`. This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The global early Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).

A model run excluding the ecological proxies results in a similar median latitudinal gradient, `r eocene_proxy_gradient$"50%"` degC, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`) degC. The inferred latitudinal gradient is somewhat steeper (`r eocene_north_south_gradient_diff$"50%"` degC, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"` degC) in the northern hemisphere, but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig. S3).)

### 2023-02-02 manuscript/supplementary_material.Rmd
results and intro updates

## S1. Convergence checks
To ensure the convergence and mixing of the chains, we inspected traceplots of all the unknown parameters of the model.

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S1. Traceplots of a subset of the unknown model parameters. The four colours correspond to the four independent model runs. a) Traceplot of A, K-A, M, B and $\\sigma$; b) Traceplot of five selected $\\mu_j$; c) Traceplot of five selected $\\sigma_j$."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS1-Traceplot.jpg", error = FALSE)
```

@@

## S3. Seperate gradients on the northen and southern hemisphere

>>>>

## S1. Convergence checks

```{r out.width='100%', echo = FALSE, fig.cap = "Fig S1. Traceplots of a subset of the unknown model parameters. The four colours correspond to the four independent model runs. a) Traceplot of A, K-A, M, B and $\\sigma$; b) Traceplot of five selected $\\mu_j$; c) Traceplot of five selected $\\sigma_j$. All traceplots display mixing of the chains, and relatively quick convergence."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS1-Traceplot.jpg", error = FALSE)
```

@@

## S3. Seperate gradients on the northern and southern hemisphere
